### 2015-03-07 paper/paper.tex
Add the start to a paper

\documentclass[acmtoplas]{acmtrans2m}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{proof}
\usepackage{twelf}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{code}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newdef{definition}[theorem]{Definition}
\newdef{remark}[theorem]{Remark}
\def\permission{}

\def\InputModeColorName{MidnightBlue}
\def\OutputModeColorName{Maroon}
\newcommand\InputMode[1]{{\color{\InputModeColorName}{#1}}}
\newcommand\OutputMode[1]{{\color{\OutputModeColorName}{#1}}}

\newcommand\naturals{\mathbb{N}}
\newcommand\type{\mathbf{type}}
\newcommand\tyvoid{\mathsf{void}}
\newcommand\tyunit{\mathsf{unit}}
\newcommand\ax{\mathsf{ax}}
\newcommand\ap{\mathsf{ap}}
\newcommand\sortval{\mathsf{val}}
\newcommand\sortexp{\mathsf{exp}}
\newcommand\sortctx{\mathsf{ctx}}
\newcommand\sorttele{\mathsf{tele}}
\newcommand\sortoexp[1]{[#1]\sortexp}
\newcommand\eval[2]{\InputMode{#1}\mathbin{\Rightarrow}\OutputMode{#2}}
\newcommand\thunk[1]{{\color{Gray}\uparrow}#1}
\newcommand\evalrule[1]{\Rightarrow_{#1}}
\newcommand\nobind[1]{{\color{Gray}!}#1}
\newcommand\bind[2]{{\color{Gray}@}[#1]#2}

\newcommand\isset[1]{\InputMode{#1}\;\mathit{set}}
\newcommand\eqset[2]{\InputMode{#1}=\InputMode{#2}\;\mathit{set}}
\newcommand\mem[2]{\InputMode{#1}\in\InputMode{#2}}
\newcommand\eqmem[3]{\InputMode{#1}=\InputMode{#2}\in\InputMode{#3}}

\newcommand\canset[1]{\InputMode{#1}\;\mathit{set}_\mathsf{v}}
\newcommand\eqcanset[2]{\InputMode{#1}=\InputMode{#2}\;\mathit{set}_\mathsf{v}}
\newcommand\canmem[2]{\InputMode{#1}\in_\mathsf{v}\InputMode{#2}}
\newcommand\eqcanmem[3]{\InputMode{#1}=\InputMode{#2}\in_\mathsf{v}\InputMode{#3}}
\newcommand\lfpi[2]{{\color{Gray}\forall\{#1\}}.\;#2}

\DeclareMathOperator{\typi}{\mathsf{pi}}
\DeclareMathOperator{\tysg}{\mathsf{sg}}
\DeclareMathOperator{\lam}{\mathsf{\lambda}}
\DeclareMathOperator{\pair}{\mathsf{pair}}
\DeclareMathOperator{\fst}{\mathsf{fst}}
\DeclareMathOperator{\snd}{\mathsf{snd}}

%% Krary Stuff
\DeclareMathOperator\isvarop{\mathsf{isvar}}
\DeclareMathOperator\boundedop{\mathsf{bounded}}
\DeclareMathOperator\orderedop{\mathsf{ordered}}
\DeclareMathOperator\precedesop{\mathsf{precedes}}

\newcommand\isvar[2]{\isvarop(\InputMode{#1};\OutputMode{#2})}
\newcommand\bounded[2]{\boundedop(\InputMode{#1};\InputMode{#2})}
\newcommand\ordered[1]{\orderedop(\InputMode{#1})}
\newcommand\precedes[2]{\precedesop(\InputMode{#1};\InputMode{#2})}
\newcommand\boundedrule[1]{\mathsf{bounded}_{#1}}
\newcommand\orderedrule[1]{\mathsf{ordered}_{#1}}
\newcommand\lessthan[2]{\InputMode{#1}<\InputMode{#2}}

\title{A Higher-Order Approach to Explicit Contexts in LF}

\author{JONATHAN M.\ STERLING}

\begin{abstract}
  There are two standard techniques for representing dependently typed calculi
  in the LF. The first is to re-use the LF contexts for the object contexts,
  which can lead to difficulties when the meaning of hypothetico-general
  judgement in the object language is stronger than in the LF; this is, for
  instance, the case in MLTT~1979, where non-trivial functionality obligations
  are incurred in the sequent judgement.

  Another technique advanced by Krary is to represent contexts explicitly, and
  then define a sequent judgement over them; this can be used to resolve the
  problem described above (and several others), but it comes at the cost of
  verbosity, and cannot be to encode \emph{telescopes}, which are contexts
  where each successive term has the previous bound variables free in it.  In
  this paper, I demonstrate a higher-order encoding of contexts and telescopes
  which can be used to faithfully encode the sequent judgement for MLTT~1979.
\end{abstract}

\category{}{}{}
\terms{Logical Frameworks}
\keywords{Contexts, telescopes}

\begin{document}

\maketitle

\section{Introduction}

To motivate this work, let us consider an encoding of MLTT~1979 in the logical
framework. The syntax contains two sorts, $\sortval$ (for canonical forms) and
$\sortexp$ (for non-canonical forms), and is defined in the following LF
signature: \footnote{LF signatures are presented in the style of inference
  rules for clarity. Object judgement forms have a \emph{mode} which expresses
  which arguments represent inputs, and which represent outputs; in our
  presentation, we color the inputs \emph{\InputMode\InputModeColorName}, and
  the outputs \emph{\OutputMode\OutputModeColorName}. Proper assignment of
modes may be used to construe a judgement as having algorithmic content.}

\begin{gather*}
  \infer{\sortval:\type}{}
  \quad
  \infer{\sortexp:\type}{}\\[6pt]
  %
  \infer{\tyvoid:\sortval}{}
  \quad
  \infer{\tyunit:\sortval}{}
  \quad
  \infer{
    \typi(A;[x]B\,x):\sortval
  }{
    A:\sortexp &
    B:\sortexp\to\sortexp
  }
  \quad
  \infer{
    \tysg(A;[x]B\,x):\sortval
  }{
    A:\sortexp &
    B:\sortexp\to\sortexp
  }\\[6pt]
  %
  \infer{\ax:\sortval}{}
  \quad
  \infer{
    \lam[x]E\,x:\sortval
  }{
    E:\sortexp\to\sortexp
  }
  \quad
  \infer{
    \pair(M;N):\sortval
  }{
    M:\sortexp &
    N:\sortexp
  } \\[6pt]
  %
  \infer{
    \thunk{M}:\sortexp
  }{
    M:\sortval
  }
  \quad
  \infer{
    \ap(M;N):\sortexp
  }{
    M:\sortexp &
    N:\sortexp
  }
  \quad
  \infer{
    \fst(M):\sortexp
  }{
    M:\sortexp
  }
  \quad
  \infer{
    \snd(M):\sortexp
  }{
    M:\sortexp
  }
\end{gather*}

Then, the language is accorded a simple big-step operational semantics via the
\framebox{$\eval{M}{N}$} judgement:

\begin{gather*}
  \infer{
    \eval{M}{N}:\type
  }{
    M:\sortexp &
    N:\sortval
  }\\[6pt]
  %
  \infer[\evalrule\uparrow]{
    \eval{\thunk{M}}{M}
  }{
  }
  \quad
  \infer[\evalrule\ap]{
    \eval{\ap(\thunk{\lam[x]E(x)};M)}{N}
  }{
    \eval{E\,M}{N}
  }\\[6pt]
  %
  \infer[\evalrule\fst]{
    \eval{\fst(\thunk{\pair(M;N)})}{M'}
  }{
    \eval{M}{M'}
  }
  \quad
  \infer[\evalrule\snd]{
    \eval{\fst(\thunk{\pair(M;N)})}{N'}
  }{
    \eval{N}{N'}
  }
\end{gather*}

\section{The Na\"ive Higher-Order Approach}

Informally, there are four basic categorical judgements in MLTT~1979:\\

\begin{tabular}{ll}
  \toprule
  Form & Pronunciation\\
  \midrule
  $\isset{A}$ & $A$ is a set\\
  $\eqset{A}{B}$ & $A$ and $B$ are equal sets\\
  $\mem{M}{A}$ & $M$ is a member of $A$\\
  $\eqmem{M}{N}{A}$ & $M$ and $N$ are equal members of $A$\\
  \bottomrule
\end{tabular}\\

\medskip\noindent The subjects of these judgements are terms of sort
$\sortexp$. However, implicit in Martin-L\"of's presentation are four prior
judgements which operate only on canonical forms (i.e. terms of sort
$\sortval$). We will therefore introduce the following LF signature to account
for this:

\begin{gather*}
  \infer{
    \canset{A}:\type
  }{
    A:\sortval
  }\quad
  \infer{
    \eqcanset{A}{B}:\type
  }{
    A:\sortval &
    B:\sortval
  }\quad
  \infer{
    \canmem{M}{A}:\type
  }{
    M:\sortval &
    A:\sortval
  }\quad
  \infer{
    \eqcanmem{M}{N}{A}:\type
  }{
    M:\sortval &
    N:\sortval &
    A:\sortval
  }\\[6pt]
  \infer{
    \isset{A}:\type
  }{
    A:\sortval
  }\quad
  \infer{
    \eqset{A}{B}:\type
  }{
    A:\sortexp &
    B:\sortexp
  }\quad
  \infer{
    \mem{M}{A}:\type
  }{
    M:\sortexp &
    A:\sortexp
  }\quad
  \infer{
    \eqmem{M}{N}{A}:\type
  }{
    M:\sortexp &
    N:\sortexp &
    A:\sortexp
  }
\end{gather*}

Then, we will give the meanings of the judgements which deal with non-canonical
forms first, anticipating the explanations of the canonical forms judgements.

\begin{gather*}
  \infer{
    \isset{A}
  }{
    \eval{A}{A'} &
    \canset{A'}
  }\quad
  \infer{
    \eqset{A}{B}
  }{
    \begin{array}{c}
      \eval{A}{A'} \\
      \eval{B}{B'}
    \end{array} &
      \eqcanset{A'}{B'}
  }\\[6pt]
  \infer{
    \mem{M}{A}
  }{
    \begin{array}{c}
      \eval{M}{M'}\\
      \eval{A}{A'}
    \end{array} &
    \canmem{M'}{A'}
  }\quad
  \infer{
    \eqmem{M}{N}{A}
  }{
    \begin{array}{c}
      \eval{M}{M'} \\
      \eval{N}{N'} \\
      \eval{A}{A'}
    \end{array} &
    \eqcanmem{M'}{N'}{A'}
  }
\end{gather*}

Following Martin-L\"of, we give an extensional equality for all types (i.e.\
two types are equal when they have the same membership and equivalence
relations); we could choose an intensional/structural equality for types, but
for our purposes here, it is not necessary.

\begin{gather*}
  \infer{
    \eqcanset{A}{B}
  }{
    \begin{array}{l}
      \lfpi{M}{\canmem{M}{A}\to\canmem{M}{B}}\\
      \lfpi{M}{\canmem{M}{B}\to\canmem{M}{A}}\\
      \lfpi{M,N}{\eqcanmem{M}{N}{A}\to\eqcanmem{M}{N}{B}}\\
      \lfpi{M,N}{\eqcanmem{M}{N}{B}\to\eqcanmem{M}{N}{A}}
    \end{array}
  }
\end{gather*}

Then, all we have to do is define the remaining canonical forms judgements with
respect to our syntax. For the base types, this is trivial:

\begin{gather*}
  \infer{
    \canset\tyvoid
  }{
  }\quad
  \infer{
    \canset\tyunit
  }{
  }\quad
  \infer{
    \canmem\ax\tyunit
  }{
  }\quad
  \infer{
    \eqcanmem\ax\ax\tyunit
  }{
  }
\end{gather*}

But for terms which involve binding, the rules will become more complicated.
Informally, Martin-L\"of gives the following formation rule for $\mathsf\Pi$
using hypothetico-general judgement:
\[
  \infer[*]{
    \canset{\mathsf{\Pi}(x\in A)B(x)}
  }{
    \isset{A} &
    \isset{B(x)}\;(\mem{x}{A})
  }
\]
So we might be tempted to do something similar in the LF, as follows:
\[
  \infer[*']{
    \canset{\typi(A;[x]B\,x)}
  }{
    \isset{A} &
    \lfpi{x}\mem{x}{A}\to\isset{B(x)}
  }
\]

This encoding, however, is not adequate, since the meaning of
hypothetico-general judgement in Martin-L\"of's informal metalangauge is not
captured by the above attempt. Recall that hypothetico-general judgement in
type theory implicitly requires \emph{functionality} of type and term families;
therefore, in order for our definition to be correct, we would have to adjust
the rule to include functionality:
\[
  \infer{
    \canset{\typi(A;[x]B\,x)}
  }{
    \isset{A} &
    \begin{array}{l}
      \lfpi{x}\mem{x}{A}\to\isset{B(x)}\\
      \lfpi{x,y}\eqmem{x}{y}{A}\to\eqset{B(x)}{B(y)}\\
    \end{array}
  }
\]

We have at least got closer, but even this is not really satisfactory; we were
unable to factor out the functionality constraints, and must be careful to
place them in all the right places in each and every rule. Contrast this with
Martin-L\"of's definition of hypothetico-general judgement, which includes
functionality from the start, allowing the other judgements to use this ``off
the shelf''.

\section{Explicit Contexts \`a la Krary}
In order to factor out functionality constraints, it will be necessary to
consider the use of contexts in the encoding, and then recast the judgements
above as being sequents, i.e.\ judgements with respect to a context. Krary's
first-order encoding of explicit contexts in the LF is essentially the
following:

\begin{gather*}
  \infer{\sortctx:\type}{}
  \quad
  \infer{\cdot:\sortctx}{}
  \quad
  \infer{
    (\Gamma,X:A):\sortctx
  }{
    \Gamma:\sortctx &
    X:\sortexp &
    A:\sortexp
  }\\[6pt]
  \infer{
    \isvar{X}{I} : \type
  }{
    X:\sortexp &
    I:\naturals
  }
  \quad
  \infer{
    \precedes{X}{Y}: \type
  }{
    X:\sortexp &
    Y:\sortexp
  }\quad
  \infer{
    \precedes{X}{Y}
  }{
    \isvar{X}{I} &
    \isvar{Y}{J} &
    \lessthan{I}{J}
  }\\[6pt]
  \infer{
    \bounded{\Gamma}{X}:\type
  }{
    \Gamma:\sortctx &
    X:\sortexp
  }
  \quad
  \infer{
    \bounded{\cdot}{X}
  }{
    \isvar{X}{I}
  }
  \quad
  \infer{
    \bounded{\Gamma,Y:A}{X}
  }{
    \precedes{Y}{X} &
    \bounded{G}{Y}
  }
  \\[6pt]
  \infer{
    \ordered\Gamma:\type
  }{
    \Gamma:\sortctx
  }
  \quad
  \infer{\ordered\cdot}{}
  \quad
  \infer{
    \ordered{\Gamma,X:A}
  }{
    \bounded{\Gamma}{X}
  }
  \\[6pt]
\end{gather*}

Note that the well-formedness constraints above do not in practice protrude
into the definitions of judgements which \emph{use} explicit contexts; they are
only used in metatheorems. This encoding is very practical for certain calculi,
but it is concerning in our case for a few reasons:

\begin{enumerate}

  \item The correctness of this encoding depends very strongly on its
    \emph{use}: this only truly encodes contexts if in every case a LF-variable
    is used for $X$.  Moreover, the various well-formedness constraints are a
    bit inelegant, as is the concept of mapping LF variables to indices.

  \item It is not possible to construct such a context on its own as a
    syntactic object of (LF-)type $\sortctx$, where variables bound earlier in
    the context may appear free later in the context. This can be done
    otherwise by constructing an object of type $\sortctx$ in an extended
    LF-context; the well-formedness of such a context is still an extrinsic
    property, though.

  \item A straightforward higher-order encoding is preferable (if possible)
    when working in the LF.

\end{enumerate}

\section{Encoding Telescopes in Higher-Order Abstract Syntax}
More precisely, it seems that what we really need is an encoding of
\emph{telescopes}, in the sense of De Bruijn. A telescope is essentially a
sequence of type assignments, where variables bound earlier may appear free
later, for instance:
\[
  \cdot,x_1:A_1, x_2:A_2(x_1), x_3:A_3(x_1,x_2), \dots, x_n:A_n(x_1,\dots,x_{n-1})
\]

To encode this in a straightforward, higher-order manner, we shall first
stipulate that each variable in the telescope will be an LF-variable. Then,
each type $A_i$ in the telescope will be a term with $i-n$ bindings (i.e.\ it
will be an $i$-ary $\lambda$-abstraction). We'll define by mutual induction a
sort for telescopes $\sorttele$, and a sort for open terms with respect to a
telescope $\sortoexp{\Psi}$:

\begin{gather*}
  \infer{\sorttele:\type}{}
  \quad
  \infer{
    \sortoexp{\Psi}:\type
  }{
    \Psi:\sorttele
  }\\[6pt]
  \infer{\cdot:\sorttele}{}
  \quad
  \infer{
    \Psi,A:\sorttele
  }{
    \Psi:\sorttele &
    A:\sortoexp{\Psi}
  }\\[6pt]
  \infer{\nobind{M}:\sortoexp{\cdot}}{}
  \quad
  \infer{
    \bind{x}{M(x)}:\sortoexp{\Psi,A}
  }{
    M:\sortexp\to\sortoexp\Psi
  }
\end{gather*}

Note that we do not put any well-typedness constraints on the arguments of the
bound term constructors; we wish to separate object syntax from object
judgements; this is particularly important for an object theory which does not
have decidable type checking.

\end{document}

### 2015-03-08 paper/paper.tex
Finish definition of the sequent judgement

@@
 \usepackage{amssymb}
+\usepackage{wasysym}
 \usepackage{booktabs}
@@
 \usepackage{code}
+\usepackage{scalerel}
 
@@
 \newcommand\sortoexp[1]{[#1]\sortexp}
+\newcommand\sortj{\mathsf{judg}}
+\newcommand\sortoj[1]{[#1]\sortj}
 \newcommand\eval[2]{\InputMode{#1}\mathbin{\Rightarrow}\OutputMode{#2}}
@@
 
+\newcommand\weaken[3]{\mathsf{wk}_{\InputMode{#1}}\,\InputMode{#2}\leadsto\OutputMode{#3}}
+\newcommand\qisset[1]{#1\;`\mathit{set}}
+\newcommand\qeqset[2]{#1=#2\;`\mathit{set}}
+\newcommand\qmem[2]{#1\;`\!\!\in#2}
+\newcommand\qeqmem[3]{#1=#2\;`\!\!\in#3}
+
 \DeclareMathOperator{\typi}{\mathsf{pi}}
@@
 \DeclareMathOperator{\snd}{\mathsf{snd}}
+
+\newcommand\sploot[2]{\InputMode{#1}\triangleright\OutputMode{#2}}
+\newcommand\refract[2]{\InputMode{#1}\ltimes\OutputMode{#2}}
+
+\newcommand\sequent[2]{\InputMode{#1}\gg\InputMode{#2}}
+\newcommand\nece[1]{\gg\InputMode{#1}}
 
@@
   \infer{
-    \typi(A;[x]B\,x):\sortval
+    \typi(A;[x]B(x)):\sortval
   }{
@@
   \infer{
-    \tysg(A;[x]B\,x):\sortval
+    \tysg(A;[x]B(x)):\sortval
   }{
@@
   \infer{
-    \lam[x]E\,x:\sortval
+    \lam[x]E(x):\sortval
   }{
@@
   }{
-    \eval{E\,M}{N}
+    \eval{E(M)}{N}
   }\\[6pt]
@@
   \infer[*']{
-    \canset{\typi(A;[x]B\,x)}
+    \canset{\typi(A;[x]B(x))}
   }{
@@
   \infer{
-    \canset{\typi(A;[x]B\,x)}
+    \canset{\typi(A;[x]B(x))}
   }{
@@
   }\\[6pt]
-  \infer{\nobind{M}:\sortoexp{\cdot}}{}
+  \infer{
+    \nobind{M}:\sortoexp{\cdot}
+  }{
+    M:\sortexp
+  }
   \quad
@@
   }{
-    M:\sortexp\to\sortoexp\Psi
+    M:\sortexp\to\sortoexp\Psi &
+    A:\sortoexp\Psi
   }
@@
 
+Closed expressions may be weakened by extraneous variables:
+\begin{gather*}
+  \infer{
+    \weaken{\Psi}{M}{M'}:\type
+  }{
+    \Psi:\sorttele &
+    M:\sortexp &
+    M':\sortoexp\Psi
+  }\\[6pt]
+  \infer{
+    \weaken{\cdot}{M}{\nobind{M}}
+  }{
+  }\quad
+  \infer{
+    \weaken{\Psi,A}{M}{\bind{x}{M'}}
+  }{
+    \weaken{\Psi}{M}{M'}
+  }
+\end{gather*}
+
+A more general definition of weakening can be given, but the simple one above
+suffices for our use-case.
+
+\section{Encoding Functional Hypothetico-General Judgement}
+Henceforth I shall use the term ``sequent judgement'' to mean a
+hypothetico-general judgement which also requires functionality. First, before
+we get to the sequent judgement, we will need the concept of an ``open
+categorical judgement''; this is a judgement which has variables from a
+telescope free in it.
+
+\subsection{Codes for Open Judgements}
+We will define \emph{codes} for such judgements and then
+interpret them into proper object judgement (LF-types) later.
+
+\begin{gather*}
+  \infer{
+    \sortoj\Psi:\type
+  }{
+    \Psi:\sorttele
+  }\\[6pt]
+  \infer{
+    \qisset{A}:\sortoj\Psi
+  }{
+    A:\sortoexp\Psi
+  }
+  \quad
+  \infer{
+    \qeqset{A}{B}:\sortoj\Psi
+  }{
+    A:\sortoexp\Psi &
+    B:\sortoexp\Psi
+  }\\[6pt]
+  \infer{
+    \qmem{M}{A}:\sortoj\Psi
+  }{
+    M:\sortoexp\Psi &
+    A:\sortoexp\Psi
+  }
+  \quad
+  \infer{
+    \qeqmem{M}{N}{A}:\sortoj\Psi
+  }{
+    M:\sortoexp\Psi &
+    N:\sortoexp\Psi &
+    A:\sortoexp\Psi
+  }
+\end{gather*}
+
+\subsection{Sploosion and Refraction}
+
+The next thing to do is to specify how each of these judgement(-codes) behaves
+operationally (i.e. by substituting its outermost variable); we will do so by
+extending the LF signature with a new judgement
+$\sploot{\mathcal{J}}{\mathcal{J}'}$ (``$\InputMode{\mathcal{J}}$ sploots into
+$\OutputMode{\mathcal{J}'}$'') and making it evident on all inputs:
+
+\begin{gather*}
+  \infer{
+    \sploot{\mathcal{J}}{[x]\mathcal{J}'(x)}:\type
+  }{
+    \mathcal{J}:\sortoj{\Psi,A} &
+    \mathcal{J}': \sortexp\to\sortoj\Psi
+  }\\[6pt]
+  \infer{
+    \sploot{
+      \qisset{\bind{x}{A(x)}}
+    }{
+      [x]\; \qisset{A(x)}
+    }
+  }{
+  }\\[6pt]
+  \infer{
+    \sploot{
+      \qeqset{\bind{x}{A(x)}}{\bind{x}{B(x)}}
+    }{
+      [x]\; \qeqset{A(x)}{B(x)}
+    }
+  }{
+  }\\[6pt]
+  \infer{
+    \sploot{
+      \qmem{\bind{x}{M(x)}}{\bind{x}{A(x)}}
+    }{
+      [x]\; \qmem{M(x)}{A(x)}
+    }
+  }{
+  }\\[6pt]
+  \infer{
+    \sploot{
+      \qeqmem{\bind{x}{M(x)}}{\bind{x}{N(x)}}{\bind{x}{A(x)}}
+    }{
+      [x]\; \qeqmem{M(x)}{N(x)}{A(x)}
+    }
+  }{
+  }
+\end{gather*}
+
+Next, we will do something similar in order to show how each of these
+judgement(-codes) behaves extensionally, by introducing another algorithmic
+judgement $\refract{\mathcal{J}}{\mathcal{J}^=}$ (``$\InputMode{\mathcal{J}}$
+refracts into $\OutputMode{\mathcal{J}^=}$''). Intuitively, this judgement
+describes how what happens when you substitute two purportedly equal
+expressions in for the judgement-code's outermost variable.
+
+\begin{gather*}
+  \infer{
+    \refract{\mathcal{J}}{[x][y]\mathcal{J}'(x)(y)}:\type
+  }{
+    \mathcal{J}:\sortoj{\Psi,A} &
+    \mathcal{J}': \sortexp\to\sortexp\to\sortoj\Psi
+  }\\[6pt]
+  \infer{
+    \refract{
+      \qisset{\bind{x}{A(x)}}
+    }{
+      [x][y]\; \qeqset{A(x)}{A(y)}
+    }
+  }{
+  }\\[6pt]
+  \infer{
+    \refract{
+      \qeqset{\bind{x}{A(x)}}{\bind{x}{B(x)}}
+    }{
+      [x][y]\; \qeqset{A(x)}{B(y)}
+    }
+  }{
+  }\\[6pt]
+  \infer{
+    \refract{
+      \qmem{\bind{x}{M(x)}}{\bind{x}{A(x)}}
+    }{
+      [x][y]\; \qeqmem{M(x)}{M(y)}{A(x)}
+    }
+  }{
+  }\\[6pt]
+  \infer{
+    \refract{
+      \qeqmem{\bind{x}{M(x)}}{\bind{x}{N(x)}}{\bind{x}{A(x)}}
+    }{
+      [x][y]\; \qeqmem{M(x)}{N(y)}{A(x)}
+    }
+  }{
+  }
+\end{gather*}
+
+\subsection{The Functional Sequent Judgement}
+
+We finally have built up enough machinery to give an immediate and elegant
+encoding to Martin-L\"of's functional sequent judgement $\mathcal{J}\ (\Psi)$;
+we will notate it $\sequent{\Psi}{\mathcal{J}}$, or $\nece{\mathcal{J}}$ when
+$\Psi$ is the empty telescope. First we will define it, and then we'll go back
+and show some corresponding examples between our system and the informal rules
+of MLTT~1979.
+
+\begin{gather*}
+  \infer{
+    \sequent{\Psi}{\mathcal{J}}:\type
+  }{
+    \Psi:\sorttele &
+    \mathcal{J}:\sortoj\Psi
+  }
+  \quad
+  \infer{
+    \nece{\mathcal{J}} \equiv \sequent{\cdot}{\mathcal{J}} : \type
+  }{
+  }
+\end{gather*}
+
+The interpretations of the judgement codes in the empty context are trivial:
+
+\begin{gather*}
+  \infer={
+    \nece{\qisset{\nobind{A}}}
+  }{
+    \isset{A}
+  }
+  \quad
+  \infer={
+    \nece{\qeqset{\nobind{A}}{\nobind{B}}}
+  }{
+    \eqset{A}{B}
+  }\\[6pt]
+  \infer={
+    \nece{\qmem{\nobind{M}}{\nobind{A}}}
+  }{
+    \mem{M}{A}
+  }
+  \quad
+  \infer={
+    \nece{\qeqmem{\nobind{M}}{\nobind{N}}{\nobind{A}}}
+  }{
+    \eqmem{M}{N}{A}
+  }
+\end{gather*}
+
+The interpretations in open contexts may be done in one fell swoop using the
+machinery that we have built up so far:
+
+\begin{gather*}
+  \infer{
+    \sequent{\Psi,A}{\mathcal{J}}
+  }{
+    \begin{array}{c}
+      \sploot{\mathcal{J}}{[x]\mathcal{J}'(x)}\\
+      \refract{\mathcal{J}}{[x][y]\mathcal{J}^=(x)(y)}\\
+      \lfpi{x,x'}{
+        \weaken{\Psi}{x}{x'}
+        \to\sequent{\Psi}{\qmem{x'}{A}}
+        \to\sequent{\Psi}{\mathcal{J}'(x)}
+      }\\
+      \lfpi{x,x',y,y'}{
+        \weaken{\Psi}{x}{x'}
+        \to\weaken{\Psi}{y}{y'}
+        \to\sequent{\Psi}{\qeqmem{x'}{y'}{A}}
+        \to\sequent{\Psi}{\mathcal{J}^=(x)(y)}
+      }\\
+    \end{array}
+  }
+\end{gather*}
+
 \end{document}

### 2015-03-08 paper/paper.tex
wordsmithing

@@
   problem described above (and several others), but it comes at the cost of
-  verbosity, and cannot be to encode \emph{telescopes}, which are contexts
-  where each successive term has the previous bound variables free in it.  In
-  this paper, I demonstrate a higher-order encoding of contexts and telescopes
-  which can be used to faithfully encode the sequent judgement for MLTT~1979.
+  verbosity, and introduces difficulties when used to encode \emph{telescopes}
+  in the LF, which are contexts where where each successive term has the
+  previous bound variables free in it.  In this paper, I demonstrate a
+  higher-order encoding of telescopes which can be used to faithfully encode
+  the functional sequent judgement for MLTT~1979.
+
 \end{abstract}
@@
 \section{Encoding Functional Hypothetico-General Judgement}
-Henceforth I shall use the term ``sequent judgement'' to mean a
+Henceforth I shall use the term ``functional sequent judgement'' to mean a
 hypothetico-general judgement which also requires functionality. First, before
-we get to the sequent judgement, we will need the concept of an ``open
+we get to the functional sequent judgement, we will need the concept of an ``open
 categorical judgement''; this is a judgement which has variables from a

### 2015-03-08 paper/paper.tex
wording

@@
   \item The correctness of this encoding depends very strongly on its
-    \emph{use}: this only truly encodes contexts if in every case a LF-variable
+    \emph{use}: this only truly encodes contexts if in every case an LF-variable
     is used for $X$.  Moreover, the various well-formedness constraints are a
-    bit inelegant, as is the concept of mapping LF variables to indices.
+    bit inelegant, as is the concept of mapping LF variables to De Bruijn indices.
 
@@
 More precisely, it seems that what we really need is an encoding of
-\emph{telescopes}, in the sense of De Bruijn. A telescope is essentially a
+\emph{telescopes}, in the sense of De Bruijn. A telescope is a
 sequence of type assignments, where variables bound earlier may appear free
-later, for instance:
+later; for instance:
 \[
@@
 stipulate that each variable in the telescope will be an LF-variable. Then,
-each type $A_i$ in the telescope will be a term with $i-n$ bindings (i.e.\ it
+each type $A_i$ in the telescope will be a term with $i-n$ LF-bindings (i.e.\ it
 will be an $i$-ary $\lambda$-abstraction). We'll define by mutual induction a
@@
 hypothetico-general judgement which also requires functionality. First, before
-we get to the functional sequent judgement, we will need the concept of an ``open
-categorical judgement''; this is a judgement which has variables from a
-telescope free in it.
+we get to the functional sequent judgement, we will need the concept of an
+``general categorical judgement''; this is a judgement which has variables from
+a telescope free in it.
 
-\subsection{Codes for Open Judgements}
+\subsection{Codes for General Judgements}
 We will define \emph{codes} for such judgements and then
-interpret them into proper object judgement (LF-types) later.
+interpret them into proper object judgements (LF-types) later.
 
@@
 operationally (i.e. by substituting its outermost variable); we will do so by
-extending the LF signature with a new judgement
+extending the LF signature with a new algorithmic judgement
 $\sploot{\mathcal{J}}{\mathcal{J}'}$ (``$\InputMode{\mathcal{J}}$ sploots into

### 2015-03-08 paper/paper.tex
Lmao, too many Ks.

@@
 
-%% Krary Stuff
+%% Crary Stuff
 \DeclareMathOperator\isvarop{\mathsf{isvar}}
@@
 
-  Another technique advanced by Krary is to represent contexts explicitly, and
+  Another technique advanced by Crary is to represent contexts explicitly, and
   then define a sequent judgement over them; this can be used to resolve the
@@
 relations); we could choose an intensional/structural equality for types, but
-for our purposes here, it is not necessary.
+this question is orthogonal to the technique being demonstrated.
 
@@
 
-\section{Explicit Contexts \`a la Krary}
+\section{Explicit Contexts \`a la Crary}
 In order to factor out functionality constraints, it will be necessary to
 consider the use of contexts in the encoding, and then recast the judgements
-above as being sequents, i.e.\ judgements with respect to a context. Krary's
+above as being sequents, i.e.\ judgements with respect to a context. Crary's
 first-order encoding of explicit contexts in the LF is essentially the

### 2015-03-08 paper/paper.tex
Add references, close #1

@@
 \begin{abstract}
+
   There are two standard techniques for representing dependently typed calculi
@@
 
-To motivate this work, let us consider an encoding of MLTT~1979 in the logical
-framework. The syntax contains two sorts, $\sortval$ (for canonical forms) and
-$\sortexp$ (for non-canonical forms), and is defined in the following LF
-signature: \footnote{LF signatures are presented in the style of inference
-  rules for clarity. Object judgement forms have a \emph{mode} which expresses
-  which arguments represent inputs, and which represent outputs; in our
-  presentation, we color the inputs \emph{\InputMode\InputModeColorName}, and
-  the outputs \emph{\OutputMode\OutputModeColorName}. Proper assignment of
-modes may be used to construe a judgement as having algorithmic content.}
+To motivate this work, let us consider an encoding of Martin-L\"of's
+extensional, polymorphic type theory (MLTT~1979) in the logical framework (see
+\cite{Martin-Lof-1979} for a full presentation). The syntax contains two sorts,
+$\sortval$ (for canonical forms) and $\sortexp$ (for non-canonical forms), and
+is defined in the following LF signature: \footnote{LF signatures are presented
+  in the style of inference rules for clarity. Object judgement forms have a
+  \emph{mode} which expresses which arguments represent inputs, and which
+  represent outputs; in our presentation, we color the inputs
+  \emph{\InputMode\InputModeColorName}, and the outputs
+  \emph{\OutputMode\OutputModeColorName}. Proper assignment of modes may be
+used to construe a judgement as having algorithmic content.}
 
@@
 
+\bibliographystyle{acmtrans}
+\nocite{*}
+\bibliography{paper}
+
 \end{document}

### 2015-03-08 paper/paper.tex
Demonstrate the encoding of pi and sigma

@@
 \newcommand\bind[2]{{\color{Gray}@}[#1]#2}
+\newcommand\tacitbind[1]{{\color{Gray}@}#1}
 
@@
 
-To motivate this work, let us consider an encoding of Martin-L\"of's
-extensional, polymorphic type theory (MLTT~1979) in the logical framework (see
-\cite{Martin-Lof-1979} for a full presentation). The syntax contains two sorts,
-$\sortval$ (for canonical forms) and $\sortexp$ (for non-canonical forms), and
-is defined in the following LF signature: \footnote{LF signatures are presented
-  in the style of inference rules for clarity. Object judgement forms have a
-  \emph{mode} which expresses which arguments represent inputs, and which
-  represent outputs; in our presentation, we color the inputs
-  \emph{\InputMode\InputModeColorName}, and the outputs
+To motivate this work, let us consider an encoding of a fragment of
+Martin-L\"of's extensional, polymorphic type theory (MLTT~1979) in the logical
+framework (see \cite{Martin-Lof-1979} for a full presentation). The syntax
+contains two sorts, $\sortval$ (for canonical forms) and $\sortexp$ (for
+non-canonical forms), and is defined in the following LF signature:
+\footnote{LF signatures are presented in the style of inference rules for
+  clarity. Object judgement forms have a \emph{mode} which expresses which
+  arguments represent inputs, and which represent outputs; in our presentation,
+  we color the inputs \emph{\InputMode\InputModeColorName}, and the outputs
   \emph{\OutputMode\OutputModeColorName}. Proper assignment of modes may be
@@
   \infer[*]{
-    \canset{\mathsf{\Pi}(x\in A)B(x)}
+    \canset{(\mathsf{\Pi}x\in A)B}
   }{
@@
 
+\section{The Definitions of the Types}
+It is now possible to give the definitions of the types; recall that we will
+only be causing judgements of the following forms to become evident:
+$\canset{A}$, $\canmem{M}{A}$, $\eqcanmem{M}{N}{A}$; all other forms of
+judgement have been fully explained. The ``spirit'' of Martin-L\"of's type
+theory lies in the concept of evaluation of closed terms to canonical form, and
+as such, defining a type means only describing how its canonical verifications
+behave.
+
+Let us recall Martin-L\"of's informal rules and we will show how they are
+encoded in the logical framework. First, the cartesian product of a family of
+sets is defined by the following rules in \cite{Martin-Lof-1979}:
+\begin{gather*}
+  \infer{
+    \isset{(\mathsf{\Pi}x\in A)B}
+  }{
+    \isset{A} &
+    \isset{B(x)}\;(\mem{x}{A})
+  }\quad
+  \infer{
+    \mem{(\lambda x)b}{(\mathsf{\Pi}x\in A)B}
+  }{
+    \mem{b}{B}\ (\mem{x}{A})
+  }\quad
+  \infer{
+    \eqmem{(\lambda x)b}{(\lambda x)d}{(\mathsf{\Pi}x\in A)B}
+  }{
+    \eqmem{b}{d}{B}\ (x\in A)
+  }
+\end{gather*}
+%
+These rules are adequately encoded in our system as follows:
+\begin{gather*}
+  \infer{
+    \canset{\typi(A;B)}
+  }{
+    \isset{A} &
+    \sequent{\cdot,A}{\qisset{\tacitbind{B}}}
+  }\quad
+  \infer{
+    \canmem{\lam(E)}{\typi(A;B)}
+  }{
+    \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
+  }\quad
+  \infer{
+    \eqcanmem{\lam(E)}{\lam(E')}{\typi(A;B)}
+  }{
+    \sequent{\cdot,A}{\eqcanmem{\tacitbind{E}}{\tacitbind{E'}}{\tacitbind{B}}}
+  }
+\end{gather*}
+
+The disjoint union of a family of sets is also easily encoded; first
+Martin-L\"of's rules:
+\begin{gather*}
+  \infer{
+    \isset{(\mathsf{\Sigma}x\in A)B}
+  }{
+    \isset{A} &
+    \isset{B(x)}\;(\mem{x}{A})
+  }\quad
+  \infer{
+    \mem{(a,b)}{(\mathsf{\Sigma}x\in A)B}
+  }{
+    \mem{a}{A} &
+    \mem{b}{[a/x]B}
+  }\quad
+  \infer{
+    \eqmem{(a,b)}{(c,d)}{(\mathsf{\Sigma}x\in A)B}
+  }{
+    \eqmem{a}{c}{a} &
+    \eqmem{b}{d}{[a/x]B}
+  }
+\end{gather*}
+%
+and their encoding:
+\begin{gather*}
+  \infer{
+    \canset{\tysg(A;B)}
+  }{
+    \isset{A} &
+    \sequent{\cdot,A}{\qisset{\tacitbind{B}}}
+  }\quad
+  \infer{
+    \canmem{\pair(M,N)}{\tysg(A;B)}
+  }{
+    \mem{M}{A} &
+    \mem{N}{B(M)}
+  }\quad
+  \infer{
+    \eqcanmem{\pair(M;N)}{\pair(M';N')}{\typi(A;B)}
+  }{
+    \eqmem{M}{M'}{A} &
+    \eqmem{N}{N'}{B(M)}
+  }
+\end{gather*}
+
+And what of the elimination rules? The entire point of the verificationist
+meaning explanations which pervade \cite{Martin-Lof-1979} is that the
+elimination rules will not be part of the definitions of the types, but rather
+rules which are admissible, justified by the meanings of the judgements and the
+definitions of the types. So, only after having propounded the definitions
+above may we entertain \emph{proving} the elimination rules in the logical
+framework.
+
+\bigskip
+
 \bibliographystyle{acmtrans}

### 2015-03-08 paper/paper.tex
corrected operational semantics
stub out admissibility lemmas
typo
[paper] Proved admissibility of function application rule

@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{case}{Case}
 \newdef{definition}[theorem]{Definition}
@@
   \infer[\evalrule\ap]{
-    \eval{\ap(\thunk{\lam[x]E(x)};M)}{N}
+    \eval{\ap(L;M)}{N}
   }{
+    \eval{L}{\lam[x]E(x)} &
     \eval{E(M)}{N}
@@
   \infer[\evalrule\fst]{
-    \eval{\fst(\thunk{\pair(M;N)})}{M'}
+    \eval{\fst(M)}{M_1'}
   }{
-    \eval{M}{M'}
+    \eval{M}{\pair(M_1;M_2)} &
+    \eval{M_1}{M_1'}
   }
@@
   \infer[\evalrule\snd]{
-    \eval{\fst(\thunk{\pair(M;N)})}{N'}
+    \eval{\snd(M)}{M_2'}
   }{
-    \eval{N}{N'}
+    \eval{M}{\pair(M_1;M_2)} &
+    \eval{M_2}{M_2'}
   }
@@
   \infer{
-    \canmem{\lam(E)}{\typi(A;B)}
+    \canmem{\lam E}{\typi(A;B)}
   }{
@@
   \infer{
-    \eqcanmem{\lam(E)}{\lam(E')}{\typi(A;B)}
+    \eqcanmem{\lam E}{\lam E'}{\typi(A;B)}
   }{
@@
 
+\subsection{Admissible Rules}
+
+\theorem{The standard application rule for the cartesian product of a family of
+sets is admissible:
+\[
+  \infer[\typi_E]{
+    \mem{\ap(M;N)}{B(N)}
+  }{
+    \begin{array}{c}
+      \mathcal{D}\\
+      \mem{M}{\thunk{\typi(A;B)}}\\
+    \end{array} &
+    \begin{array}{c}
+      \mathcal{E}\\
+      \mem{N}{A}
+    \end{array}
+  }
+\]
+}
+\begin{proof}
+  By induction on the derivation $\mathcal{D}$, which must have the following canonical form:
+  \[
+     \infer[\mathcal{D}]{
+       \mem{M}{\thunk{\typi(A;B)}}
+     }{
+       \begin{array}{c}
+         \mathcal{D}' \\
+         \eval{M}{\lam E}
+       \end{array} &
+       \infer{
+         \canmem{\lam E}{\typi(A;B)}
+       }{
+         \infer{
+           \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
+         }{
+           \infer{
+             \lfpi{x,x'}{
+               \weaken{\cdot}{x}{x'}
+               \to\sequent{\cdot}{\qmem{x'}{A}}
+               \to\sequent{\cdot}{\qmem{\nobind{E(x)}}{\nobind{B(x)}}}
+             }
+           }{
+             \begin{array}{c}
+               \mathcal{D}''\\
+               \lfpi{x}{\mem{x}{A} \to \mem{E(x)}{B(x)}}
+             \end{array}
+           } &
+           \dots
+         }
+       }
+     }
+  \]
+
+  Admissibility follows from $\mathcal{D}'$, $\mathcal{D}''$, $\mathcal{E}$;
+  the derivation $\mathcal{D''}(N)(\mathcal{E})$ must have a canonical form as follows:
+  \[
+    \infer{
+      \mem{E(N)}{B(N)}
+    }{
+      \begin{array}{c}
+        \mathcal{F}\\
+        \eval{E(N)}{X}
+      \end{array} &
+      \begin{array}{c}
+        \mathcal{F}'\\
+        \eval{B(N)}{C}
+      \end{array} &
+      \begin{array}{c}
+        \mathcal{F}''\\
+        \canmem{X}{C}
+      \end{array}
+    }
+  \]
+  Whence we may construct the goal:
+  \[
+    \infer{
+      \mem{\ap(M;N)}{B(N)}
+    }{
+      \infer{
+        \eval{\ap(M;N)}{X}
+      }{
+        \infer[\mathcal{D}']{\eval{M}{\lam E}}{} &
+        \infer[\mathcal{F}]{\eval{E(N)}{X}}{}
+      }&
+      \infer[\mathcal{F}']{\eval{B(N)}{C}}{}&
+      \infer[\mathcal{F}'']{\canmem{X}{C}}{}&
+    }
+  \]
+\end{proof}
+
+\theorem{The two standard elimination rules for the disjoint union of a family of sets are admissible:
+\[
+  \infer[\tysg_{E_1}]{
+    \mem{\fst(M)}{A}
+  }{
+    \mem{M}{\thunk{\tysg(A;[x]B(x))}} &
+  }\quad
+  \infer[\tysg_{E_2}]{
+    \mem{\snd(M)}{B(\fst(M))}
+  }{
+    \mem{M}{\thunk{\tysg(A;[x]B(x))}} &
+  }
+\]
+}
+\begin{proof}
+\end{proof}
+
 \bigskip

### 2015-03-08 paper/paper.tex
tweaks in admissibility proof formatting
typo fix
s/proving/justifying
typos

@@
 stipulate that each variable in the telescope will be an LF-variable. Then,
-each type $A_i$ in the telescope will be a term with $i-n$ LF-bindings (i.e.\ it
-will be an $i$-ary $\lambda$-abstraction). We'll define by mutual induction a
+each type $A_i$ in the telescope will be a term with $i-1$ LF-bindings (i.e.\ it
+will be an $(i-1)$-ary $\lambda$-abstraction). We'll define by mutual induction a
 sort for telescopes $\sorttele$, and a sort for open terms with respect to a
@@
   }{
-    \eqmem{a}{c}{a} &
+    \eqmem{a}{c}{A} &
     \eqmem{b}{d}{[a/x]B}
@@
 definitions of the types. So, only after having propounded the definitions
-above may we entertain \emph{proving} the elimination rules in the logical
+above may we entertain \emph{justifying} the elimination rules in the logical
 framework.
@@
   \[
-    \infer{
+    \infer[\mathcal{D''}(N)(\mathcal{E})]{
       \mem{E(N)}{B(N)}
@@
       }{
-        \infer[\mathcal{D}']{\eval{M}{\lam E}}{} &
-        \infer[\mathcal{F}]{\eval{E(N)}{X}}{}
+        \begin{array}{c}
+          \mathcal{D}'\\
+          \eval{M}{\lam E}
+        \end{array} &
+        \begin{array}{c}
+          \mathcal{F}\\
+          \eval{E(N)}{X}
+        \end{array}
       }&
-      \infer[\mathcal{F}']{\eval{B(N)}{C}}{}&
-      \infer[\mathcal{F}'']{\canmem{X}{C}}{}&
+      \begin{array}{c}
+        \mathcal{F}'\\
+        \eval{B(N)}{C}
+      \end{array}&
+      \begin{array}{c}
+        \mathcal{F}''\\
+        \canmem{X}{C}
+      \end{array}
     }
@@
   }{
-    \mem{M}{\thunk{\tysg(A;[x]B(x))}} &
+    \mem{M}{\thunk{\tysg(A;B)}} &
   }\quad
@@
   }{
-    \mem{M}{\thunk{\tysg(A;[x]B(x))}} &
+    \mem{M}{\thunk{\tysg(A;B)}} &
   }

### 2015-03-08 paper/paper.tex
repair contradiction in terms

@@
 hypothetico-general judgement which also requires functionality. First, before
-we get to the functional sequent judgement, we will need the concept of an
-``general categorical judgement''; this is a judgement which has variables from
+we get to the full sequent judgement, we will need the concept of an
+``general judgement''; this is a judgement which has variables from
 a telescope free in it.

### 2015-03-08 paper/paper.tex
hardening proof of theorem 6.1
Add \dlabel macro for labelling derivations
proved sigma elims; will circle back after I try it in twelf

@@
 \newcommand\weaken[3]{\mathsf{wk}_{\InputMode{#1}}\,\InputMode{#2}\leadsto\OutputMode{#3}}
+\newcommand\weakenrulenil{\mathsf{wk}_\mathsf{nil}}
+\newcommand\weakenrulesnoc{\mathsf{wk}_\mathsf{snoc}}
 \newcommand\qisset[1]{#1\;`\mathit{set}}
@@
 \newcommand\nece[1]{\gg\InputMode{#1}}
+
+\newcommand\sequentruleset{\gg_\mathsf{set}}
+\newcommand\sequentruleeqset{\gg_{\mathsf{set}_=}}
+\newcommand\sequentrulemem{\gg_\in}
+\newcommand\sequentruleeqmem{\gg_{\in_=}}
+\newcommand\dname[1]{\mathcal{#1}}
+\newcommand\dlabel[2]{\begin{array}{c}#1\\#2\end{array}}
 
@@
 extending the LF signature with a new algorithmic judgement
-$\sploot{\mathcal{J}}{\mathcal{J}'}$ (``$\InputMode{\mathcal{J}}$ sploots into
-$\OutputMode{\mathcal{J}'}$'') and making it evident on all inputs:
-
-\begin{gather*}
-  \infer{
-    \sploot{\mathcal{J}}{[x]\mathcal{J}'(x)}:\type
-  }{
-    \mathcal{J}:\sortoj{\Psi,A} &
-    \mathcal{J}': \sortexp\to\sortoj\Psi
+$\sploot{\dname{J}}{\dname{J}'}$ (``$\InputMode{\dname{J}}$ sploots into
+$\OutputMode{\dname{J}'}$'') and making it evident on all inputs:
+
+\begin{gather*}
+  \infer{
+    \sploot{\dname{J}}{[x]\dname{J}'(x)}:\type
+  }{
+    \dname{J}:\sortoj{\Psi,A} &
+    \dname{J}': \sortexp\to\sortoj\Psi
   }\\[6pt]
@@
 judgement(-codes) behaves extensionally, by introducing another algorithmic
-judgement $\refract{\mathcal{J}}{\mathcal{J}^=}$ (``$\InputMode{\mathcal{J}}$
-refracts into $\OutputMode{\mathcal{J}^=}$''). Intuitively, this judgement
+judgement $\refract{\dname{J}}{\dname{J}^=}$ (``$\InputMode{\dname{J}}$
+refracts into $\OutputMode{\dname{J}^=}$''). Intuitively, this judgement
 describes how what happens when you substitute two purportedly equal
@@
   \infer{
-    \refract{\mathcal{J}}{[x][y]\mathcal{J}'(x)(y)}:\type
-  }{
-    \mathcal{J}:\sortoj{\Psi,A} &
-    \mathcal{J}': \sortexp\to\sortexp\to\sortoj\Psi
+    \refract{\dname{J}}{[x][y]\dname{J}'(x)(y)}:\type
+  }{
+    \dname{J}:\sortoj{\Psi,A} &
+    \dname{J}': \sortexp\to\sortexp\to\sortoj\Psi
   }\\[6pt]
@@
 We finally have built up enough machinery to give an immediate and elegant
-encoding to Martin-L\"of's functional sequent judgement $\mathcal{J}\ (\Psi)$;
-we will notate it $\sequent{\Psi}{\mathcal{J}}$, or $\nece{\mathcal{J}}$ when
+encoding to Martin-L\"of's functional sequent judgement $\dname{J}\ (\Psi)$;
+we will notate it $\sequent{\Psi}{\dname{J}}$, or $\nece{\dname{J}}$ when
 $\Psi$ is the empty telescope. First we will define it, and then we'll go back
@@
   \infer{
-    \sequent{\Psi}{\mathcal{J}}:\type
+    \sequent{\Psi}{\dname{J}}:\type
   }{
     \Psi:\sorttele &
-    \mathcal{J}:\sortoj\Psi
-  }
-  \quad
-  \infer{
-    \nece{\mathcal{J}} \equiv \sequent{\cdot}{\mathcal{J}} : \type
+    \dname{J}:\sortoj\Psi
+  }
+  \quad
+  \infer{
+    \nece{\dname{J}} \equiv \sequent{\cdot}{\dname{J}} : \type
   }{
@@
 \begin{gather*}
-  \infer={
+  \infer=[\sequentruleset]{
     \nece{\qisset{\nobind{A}}}
@@
   \quad
-  \infer={
+  \infer=[\sequentruleeqset]{
     \nece{\qeqset{\nobind{A}}{\nobind{B}}}
@@
   }\\[6pt]
-  \infer={
+  \infer=[\sequentrulemem]{
     \nece{\qmem{\nobind{M}}{\nobind{A}}}
@@
   \quad
-  \infer={
+  \infer=[\sequentruleeqmem]{
     \nece{\qeqmem{\nobind{M}}{\nobind{N}}{\nobind{A}}}
@@
   \infer{
-    \sequent{\Psi,A}{\mathcal{J}}
+    \sequent{\Psi,A}{\dname{J}}
   }{
     \begin{array}{c}
-      \sploot{\mathcal{J}}{[x]\mathcal{J}'(x)}\\
-      \refract{\mathcal{J}}{[x][y]\mathcal{J}^=(x)(y)}\\
+      \sploot{\dname{J}}{[x]\dname{J}'(x)}\\
+      \refract{\dname{J}}{[x][y]\dname{J}^=(x)(y)}\\
       \lfpi{x,x'}{
@@
         \to\sequent{\Psi}{\qmem{x'}{A}}
-        \to\sequent{\Psi}{\mathcal{J}'(x)}
+        \to\sequent{\Psi}{\dname{J}'(x)}
       }\\
@@
         \to\sequent{\Psi}{\qeqmem{x'}{y'}{A}}
-        \to\sequent{\Psi}{\mathcal{J}^=(x)(y)}
+        \to\sequent{\Psi}{\dname{J}^=(x)(y)}
       }\\
@@
   }{
-    \begin{array}{c}
-      \mathcal{D}\\
-      \mem{M}{\thunk{\typi(A;B)}}\\
-    \end{array} &
-    \begin{array}{c}
-      \mathcal{E}\\
+    \dlabel{\dname D}{
+      \mem{M}{\thunk{\typi(A;B)}}
+    } &
+   \dlabel{\dname E}{
       \mem{N}{A}
-    \end{array}
+    }
   }
@@
 \begin{proof}
-  By induction on the derivation $\mathcal{D}$, which must have the following canonical form:
+  By induction on the derivation $\dname{D}$, which must have the following canonical form:
   \[
-     \infer[\mathcal{D}]{
+     \infer[\dname{D}]{
        \mem{M}{\thunk{\typi(A;B)}}
      }{
-       \begin{array}{c}
-         \mathcal{D}' \\
-         \eval{M}{\lam E}
-       \end{array} &
+       \dlabel{\dname D}{\eval{M}{\lam E}} &
        \infer{
@@
          }{
-           \infer{
+           \dlabel{\dname D''}{
              \lfpi{x,x'}{
@@
              }
-           }{
-             \begin{array}{c}
-               \mathcal{D}''\\
-               \lfpi{x}{\mem{x}{A} \to \mem{E(x)}{B(x)}}
-             \end{array}
            } &
-           \dots
+           \dlabel{}{\dots}
          }
@@
 
-  Admissibility follows from $\mathcal{D}'$, $\mathcal{D}''$, $\mathcal{E}$;
-  the derivation $\mathcal{D''}(N)(\mathcal{E})$ must have a canonical form as follows:
+  Admissibility follows from $\dname{D}'$, $\dname{D}''$, $\dname{E}$;
+  let $\dname{F} \equiv
+  \dname{D}''(N,\nobind{N},\weakenrulenil,\sequentrulemem\dname{E})$, which
+  must be a canonical derivation as follows:
   \[
-    \infer[\mathcal{D''}(N)(\mathcal{E})]{
+    \infer[\dname{F}]{
+      \sequent{\cdot}{\qmem{\nobind{E(N)}}{\nobind{B(N)}}}
+    }{
+      \infer{
         \mem{E(N)}{B(N)}
       }{
-      \begin{array}{c}
-        \mathcal{F}\\
-        \eval{E(N)}{X}
-      \end{array} &
-      \begin{array}{c}
-        \mathcal{F}'\\
-        \eval{B(N)}{C}
-      \end{array} &
-      \begin{array}{c}
-        \mathcal{F}''\\
-        \canmem{X}{C}
-      \end{array}
+        \dlabel{\dname F'}{\eval{E(N)}{X}} &
+        \dlabel{\dname F''}{\eval{B(N)}{C}} &
+        \dlabel{\dname F'''}{\canmem{X}{C}}
+      }
     }
@@
   Whence we may construct the goal:
+  \medskip
+
+  \framebox{
+    \begin{minipage}{0.9\textwidth}
       \[
@@
           }{
-        \begin{array}{c}
-          \mathcal{D}'\\
-          \eval{M}{\lam E}
-        \end{array} &
-        \begin{array}{c}
-          \mathcal{F}\\
-          \eval{E(N)}{X}
-        \end{array}
+            \dlabel{\dname D'}{\eval{M}{\lam E}} &
+            \dlabel{\dname F'}{\eval{E(N)}{X}}
           }&
-      \begin{array}{c}
-        \mathcal{F}'\\
-        \eval{B(N)}{C}
-      \end{array}&
-      \begin{array}{c}
-        \mathcal{F}''\\
-        \canmem{X}{C}
-      \end{array}
+          \dlabel{\dname F''}{\eval{B(N)}{C}} &
+          \dlabel{\dname F'''}{\canmem{X}{C}}
         }
       \]
+    \end{minipage}
+  }
 \end{proof}
@@
   }{
-    \mem{M}{\thunk{\tysg(A;B)}} &
+    \dlabel{\dname D}{\mem{M}{\thunk{\tysg(A;B)}}}
   }\quad
@@
   }{
-    \mem{M}{\thunk{\tysg(A;B)}} &
+    \dlabel{\dname D}{\mem{M}{\thunk{\tysg(A;B)}}}
   }
@@
 \begin{proof}
+
+  The derivations of the elimination rules are constructed by induction on
+  $\dname{D}$, which must be a canonical derivation of the following form:
+  \[
+    \infer[\dname{D}]{
+      \mem{M}{\thunk{\tysg(A;B)}}
+    }{
+      \dlabel{\dname D_1}{
+        \eval{M}{\pair(M_1,M_2)}
+      } &
+      \infer{
+        \canmem{\pair(M_1,M_2)}{\tysg(A;B)}
+      }{
+        \infer{
+          \mem{M_1}{A}
+        }{
+          \dlabel{\dname D_2}{\eval{M_1}{X}} &
+          \dlabel{\dname D_3}{\eval{A}{A'}} &
+          \dlabel{\dname D_4}{\canmem{X}{A'}}
+        }&
+        \infer[\dname D_5]{
+          \mem{M_2}{B(M_1)}
+        }{
+          \dlabel{\dname D_6}{\eval{M_2}{Y}} &
+          \dlabel{\dname D_7}{\eval{B(M_1)}{C}} &
+          \dlabel{\dname D_8}{\canmem{Y}{C}}
+        }
+      }
+    }
+  \]
+
+  The two elimination rules are derived as follows:
+  \medskip
+
+  \framebox{
+    \begin{minipage}{0.9\textwidth}
+      \begin{gather*}
+        \infer[\tysg_{E_1}]{
+          \mem{\fst(M)}{A}
+        }{
+          \infer{
+            \eval{\fst(M)}{X}
+          }{
+            \dlabel{\dname D_1}{\eval{M}{\pair(M_1,M_2)}} &
+            \dlabel{\dname D_2}{\eval{M_1}{X}}
+          } &
+          \dlabel{\dname D_3}{\eval{A}{A'}} &
+          \dlabel{\dname D_4}{\canmem{X}{A'}}
+        }\\[12pt]
+        \infer[\tysg_{E_2}]{
+          \mem{\snd(M)}{B(\fst(M))}
+        }{
+          \infer{
+            \eval{\snd(M)}{Y}
+          }{
+            \dlabel{\dname D_1}{\eval{M}{\pair(M_1,M_2)}} &
+            \dlabel{\dname D_6}{\eval{M_2}{Y}}
+          }&
+          \dlabel{\dname D_7}{\eval{B(\fst(M))}{C}} & % TODO: A little fishy
+          \dlabel{\dname D_8}{\canmem{Y}{C}}
+        }
+      \end{gather*}
+    \end{minipage}
+  }
 \end{proof}

### 2015-03-28 paper/paper.tex
remove sketchy proofs for now
wordsmithing, conclusion

@@
 
-  \item The correctness of this encoding depends very strongly on its
+  \item The adequacy of this encoding depends very strongly on its
     \emph{use}: this only truly encodes contexts if in every case an LF-variable
@@
 More precisely, it seems that what we really need is an encoding of
-\emph{telescopes}, in the sense of De Bruijn. A telescope is a
-sequence of type assignments, where variables bound earlier may appear free
-later; for instance:
+\emph{telescopes}, in the sense of De
+Bruijn\cite{Bruijn91telescopicmappings}. A telescope is a sequence of
+type assignments, where variables bound earlier may appear free later;
+for instance:
 \[
@@
 extending the LF signature with a new algorithmic judgement
-$\sploot{\dname{J}}{\dname{J}'}$ (``$\InputMode{\dname{J}}$ sploots into
-$\OutputMode{\dname{J}'}$'') and making it evident on all inputs:
+$\sploot{\dname{J}}{\dname{J}'}$ and making it evident on all inputs:
 
@@
 judgement(-codes) behaves extensionally, by introducing another algorithmic
-judgement $\refract{\dname{J}}{\dname{J}^=}$ (``$\InputMode{\dname{J}}$
-refracts into $\OutputMode{\dname{J}^=}$''). Intuitively, this judgement
+judgement $\refract{\dname{J}}{\dname{J}^=}$. Intuitively, this judgement
 describes how what happens when you substitute two purportedly equal
@@
 \end{gather*}
+
+The judgement $\sploot{\dname{J}}{\dname{J}'}$ may be considered as
+defining the behavior of $\dname{J}$ with one hole punched in it;
+$[x]\;\dname{J}(x)$ could be called the ``one-hole application of
+$\dname{J}$''. Likewise, $\refract{\dname{J}}{\dname{J}^=}$ may be
+thought of as defining the behavior $\dname{J}$ with two hole punched
+in it, and $[x][y]\; \dname{J}^=(x,y)$ could be called the ``two-hole
+application of $\dname{J}$''.
 
@@
 
-\subsection{Admissible Rules}
-
-\theorem{The standard application rule for the cartesian product of a family of
+\subsection{Validity of Elimination Rules}
+
+Here, we demonstrate the validity of the application rule for $\typi$
+sets; the validity of further elimination rules is left for future
+work.
+
+\begin{theorem}
+The standard application rule for the cartesian product of a family of
 sets is admissible:
@@
 \]
-}
+\end{theorem}
+
 \begin{proof}
@@
 
-\theorem{The two standard elimination rules for the disjoint union of a family of sets are admissible:
-\[
-  \infer[\tysg_{E_1}]{
-    \mem{\fst(M)}{A}
-  }{
-    \dlabel{\dname D}{\mem{M}{\thunk{\tysg(A;B)}}}
-  }\quad
-  \infer[\tysg_{E_2}]{
-    \mem{\snd(M)}{B(\fst(M))}
-  }{
-    \dlabel{\dname D}{\mem{M}{\thunk{\tysg(A;B)}}}
-  }
-\]
-}
-\begin{proof}
-
-  The derivations of the elimination rules are constructed by induction on
-  $\dname{D}$, which must be a canonical derivation of the following form:
-  \[
-    \infer[\dname{D}]{
-      \mem{M}{\thunk{\tysg(A;B)}}
-    }{
-      \dlabel{\dname D_1}{
-        \eval{M}{\pair(M_1,M_2)}
-      } &
-      \infer{
-        \canmem{\pair(M_1,M_2)}{\tysg(A;B)}
-      }{
-        \infer{
-          \mem{M_1}{A}
-        }{
-          \dlabel{\dname D_2}{\eval{M_1}{X}} &
-          \dlabel{\dname D_3}{\eval{A}{A'}} &
-          \dlabel{\dname D_4}{\canmem{X}{A'}}
-        }&
-        \infer[\dname D_5]{
-          \mem{M_2}{B(M_1)}
-        }{
-          \dlabel{\dname D_6}{\eval{M_2}{Y}} &
-          \dlabel{\dname D_7}{\eval{B(M_1)}{C}} &
-          \dlabel{\dname D_8}{\canmem{Y}{C}}
-        }
-      }
-    }
-  \]
-
-  The two elimination rules are derived as follows:
-  \medskip
-
-  \framebox{
-    \begin{minipage}{0.9\textwidth}
-      \begin{gather*}
-        \infer[\tysg_{E_1}]{
-          \mem{\fst(M)}{A}
-        }{
-          \infer{
-            \eval{\fst(M)}{X}
-          }{
-            \dlabel{\dname D_1}{\eval{M}{\pair(M_1,M_2)}} &
-            \dlabel{\dname D_2}{\eval{M_1}{X}}
-          } &
-          \dlabel{\dname D_3}{\eval{A}{A'}} &
-          \dlabel{\dname D_4}{\canmem{X}{A'}}
-        }\\[12pt]
-        \infer[\tysg_{E_2}]{
-          \mem{\snd(M)}{B(\fst(M))}
-        }{
-          \infer{
-            \eval{\snd(M)}{Y}
-          }{
-            \dlabel{\dname D_1}{\eval{M}{\pair(M_1,M_2)}} &
-            \dlabel{\dname D_6}{\eval{M_2}{Y}}
-          }&
-          \dlabel{\dname D_7}{\eval{B(\fst(M))}{C}} & % TODO: A little fishy
-          \dlabel{\dname D_8}{\canmem{Y}{C}}
-        }
-      \end{gather*}
-    \end{minipage}
-  }
-\end{proof}
+\section{Conclusion}
+
+A rich, higher-order encoding for telescopes addresses several issues
+which come up when encoding Martin-L\"of's extensional type theory
+into the LF.
+
+First, functionality premises no longer need be directly included in
+every rule which uses hypothetico-general judgement. These are
+factored out into a separate judgement which has a compositional
+definition (the functional sequent judgement). Because of this, the
+rules in the LF will look very much like how they did in
+Mpartin-L\"of's informal presentation.
+
+Secondly, the core judgements of type theory (i.e.\ typehood, equality
+of types, membership, and equality of members) are defined first in
+their atomic/categorical form, and then separately they are explained
+with respect to a telescope, both directly and extensionally. Unlike
+in Martin-L\"of's presentation, the sequent judgement is, then, not
+defined separately at each categorical judgement, but is rather
+explained once and for all in terms of the novel
+$\sploot{\dname{J}}{\dname{J}'}$ and
+$\refract{\dname{J}}{\dname{J}^=}$ judgements.
 

### 2015-03-29 paper/paper.tex
header fix

@@
 To motivate this work, let us consider an encoding of a fragment of
-Martin-L\"of's extensional, polymorphic type theory (MLTT~1979) in the logical
-framework (see \cite{Martin-Lof-1979} for a full presentation). The syntax
-contains two sorts, $\sortval$ (for canonical forms) and $\sortexp$ (for
-non-canonical forms), and is defined in the following LF signature:
-\footnote{LF signatures are presented in the style of inference rules for
-  clarity. Object judgement forms have a \emph{mode} which expresses which
-  arguments represent inputs, and which represent outputs; in our presentation,
-  we color the inputs \emph{\InputMode\InputModeColorName}, and the outputs
-  \emph{\OutputMode\OutputModeColorName}. Proper assignment of modes may be
-used to construe a judgement as having algorithmic content.}
+Martin-L\"of's extensional, polymorphic type
+theory\cite{Martin-Lof-1979} in the logical framework. The syntax
+contains two sorts, $\sortval$ (for canonical forms) and $\sortexp$
+(for non-canonical forms), and is defined in the following LF
+signature:\footnote{LF signatures are presented in the style of
+inference rules for clarity. Object judgement forms have a \emph{mode}
+which expresses which arguments represent inputs, and which represent
+outputs; in our presentation, we color the inputs
+\emph{\InputMode\InputModeColorName}, and the outputs
+\emph{\OutputMode\OutputModeColorName}. Proper assignment of modes may
+be used to construe a judgement as having algorithmic content.}
 
@@
 
-\subsection{Sploosion and Refraction}
+\subsection{Extensionality}
 
@@
 rules in the LF will look very much like how they did in
-Mpartin-L\"of's informal presentation.
+Martin-L\"of's informal presentation.
 

### 2015-03-30 paper/paper.tex
typo

@@
 hypothetico-general judgement which also requires functionality. First, before
-we get to the full sequent judgement, we will need the concept of an
+we get to the full sequent judgement, we will need the concept of a
 ``general judgement''; this is a judgement which has variables from

### 2015-03-30 paper/paper.tex
typos

@@
   verbosity, and introduces difficulties when used to encode \emph{telescopes}
-  in the LF, which are contexts where where each successive term has the
+  in the LF, which are contexts where each successive term has the
   previous bound variables free in it.  In this paper, I demonstrate a
@@
 Martin-L\"of's extensional, polymorphic type
-theory\cite{Martin-Lof-1979} in the logical framework. The syntax
+theory \cite{Martin-Lof-1979} in the logical framework. The syntax
 contains two sorts, $\sortval$ (for canonical forms) and $\sortexp$
@@
 \emph{telescopes}, in the sense of De
-Bruijn\cite{Bruijn91telescopicmappings}. A telescope is a sequence of
+Bruijn \cite{Bruijn91telescopicmappings}. A telescope is a sequence of
 type assignments, where variables bound earlier may appear free later;
@@
 judgement $\refract{\dname{J}}{\dname{J}^=}$. Intuitively, this judgement
-describes how what happens when you substitute two purportedly equal
+describes what happens when you substitute two purportedly equal
 expressions in for the judgement-code's outermost variable.
@@
 
-The judgement $\sploot{\dname{J}}{\dname{J}'}$ may be considered as
-defining the behavior of $\dname{J}$ with one hole punched in it;
-$[x]\;\dname{J}(x)$ could be called the ``one-hole application of
-$\dname{J}$''. Likewise, $\refract{\dname{J}}{\dname{J}^=}$ may be
-thought of as defining the behavior $\dname{J}$ with two hole punched
-in it, and $[x][y]\; \dname{J}^=(x,y)$ could be called the ``two-hole
-application of $\dname{J}$''.
+The judgement $\sploot{\dname{J}}{\dname{J}'}$ may be considered as defining
+the behavior of $\dname{J}$ with one hole punched in it; $[x]\;\dname{J}(x)$
+could be called the ``one-hole application of $\dname{J}$''. Likewise,
+$\refract{\dname{J}}{\dname{J}^=}$ may be thought of as defining the behavior
+$\dname{J}$ with two holes punched in it, and $[x][y]\; \dname{J}^=(x,y)$ could
+be called the ``two-hole application of $\dname{J}$''.
 
